---
title: "IEAP-Series02-Rstudio: Correlation and statistical tests\n"
author: "Morgan VIROLAN"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 Correlations and linear regression
In this series of exercises, we will explore correlations and linear regression, using a dataset from a real experiment.

## 1.1 The problem
We want to know if there is a relationship between the nonuse of the proximal part of the upper limb (PANU) and the nonuse of the shoulder (SANU) or the elbow (EENU).

## 1.2 The data
As we will see, the data is far from perfect, but this is the reality of experimental data, especially in the case of clinical data.

The file "NonUse.csv" contains measures of upper extremity nonuse:

- PANU: Proximal Arm Non Use
- SANU: Shoulder Antepulsion Non Use
- EENU: Elbow Extension Non Use

In the notebook, we'll make a clear description of the experiment, the variables and the data, providing the units of each variable, and the range of possible values.


```{r load-data and create descriptive table}
# loading knit for table generation down the line
library(knitr)
# loading the csv file
nonuse <- read.csv("data/test_data/NonUse.csv" , header = TRUE)
summary(nonuse) 


tbl_PaSaEe <- data.frame(
  Name = c("PANU", "SANU", "EENU"),
  Description = c(
    "Proximal Arm Non Use",
    "Shoulder Antepulsion Non Use",
    "Elbow Extension Non Use"
  ),
  unit = "%",
  range = "0-100"
)

kable(tbl_PaSaEe, caption = "Table 1: Description of the variables in NonUse.csv")
```

## 1.3 Correlation analysis

**Is PANU correlated with SANU and/or EENU?**

We use Spearman’s rank correlation because the data are clinical, not guaranteed to follow a normal distribution, and the relationships may be monotonic but not strictly linear.

```{r correlation tests}
cor.test(nonuse$PANU, nonuse$SANU, method = "spearman")
cor.test(nonuse$PANU, nonuse$EENU, method = "spearman")
```
- PANU and SANU show a weak but significant positive correlation (ρ = 0.22, p = 0.002, both values rounded up).
- PANU and EENU show a moderate and significant positive correlation (ρ = 0.39, p < 0.001, both values rounded up).

This means that higher proximal arm nonuse is associated with higher nonuse of the shoulder and elbow, with a stronger relationship at the elbow level.

**Visualize the PANU-SANU relationship (and the PANU-EENU relationship)**

```{r scatter plots visualisation of the relationships}
library(ggplot2)

plot_scatter <- function(df, xvar, yvar, xlabel, ylabel) {
  ggplot(df, aes_string(x = xvar, y = yvar)) +
    geom_point(na.rm = TRUE) + # added to remove the warning on missing data
    labs(title = paste("Scatter plot of", xlabel, "vs", ylabel),
         x = xlabel,
         y = ylabel) +
    coord_equal() + # auto scale, to have the same scale on both axis
    theme_minimal()
}

plot_scatter(nonuse, "PANU", "SANU", "PANU (%)", "SANU (%)")
plot_scatter(nonuse, "PANU", "EENU", "PANU (%)", "EENU (%)")

```
**Perform a linear regression for PANU-SANU (and for PANU-EENU)**
We do not predict a linear relationship between the variables but we do as we are asked. 

```{r plot-sanu-lm}
library (ggplot2)

ggplot(nonuse, aes(x = SANU, y = PANU)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Scatter plot of PANU vs SANU with regression line",
       x = "SANU (%)",
       y = "PANU (%)") +
  coord_equal() + # auto scale, to have the same scale on both axis
  theme_minimal()

ggplot(nonuse, aes(x = EENU, y = PANU)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Scatter plot of PANU vs EENU with regression line",
       x = "EENU (%)",
       y = "PANU (%)") +
  coord_equal() + # auto scale, to have the same scale on both axis
  theme_minimal()
```


**Give the regression equations: PANU = a * SANU + b and PANU = a * EENU + b**

```{r}
lm_sanu <- lm(PANU ~ SANU, data = nonuse)
summary(lm_sanu)
coef(lm_sanu)
slope <- coef(lm_sanu)[2]
intercept <- coef(lm_sanu)[1]
cat("The regression equation is: PANU =", round(slope,3), "* SANU +", round(intercept,2), "\n")

lm_eenu <- lm(PANU ~ EENU, data = nonuse)
summary(lm_eenu)
coef(lm_eenu)
slope <- coef(lm_eenu)[2]
intercept <- coef(lm_eenu)[1]
cat("The regression equation is: PANU =", round(slope,3), "* EENU +", round(intercept,2), "\n")

```

The linear model PANU ~ SANU yielded the equation:

$$
PANU = 0.101 \times SANU + 8.87
$$

This relationship was not statistically significant* (p = 0.171) and explained less than 1% of the variance (R² = 0.009), indicating very weak predictive power.  

The model PANU ~ EENU yielded the equation:

$$
PANU = 0.399 \times EENU + 5.67
$$

This relationship was statistically significant* (p < 0.001) and explained about 22% of the variance (R² = 0.225), suggesting a modest but meaningful association between proximal arm nonuse and elbow extension nonuse.


**Summarize your results as you would do in your master’s thesis report, or as in a scientific article (e.g., What sentence will you write in the article explaining your result?).**

The model PANU ~ EENU was statistically significant (p < 0.001), with a slope of 0.399. About 22% of the variance in PANU was explained by EENU, indicating a modest but meaningful association. Finally, the analysis is correlational and does not imply causality.

**What are the limits of your analysis?**

These analyses are limited by the large number of missing values (223 observations removed), the weak explanatory power of the PANU ~ SANU model, and the assumptions of linear regression (linearity, normally distributed residuals, homoscedasticity), which may not be fully met. 

# 2 Statistical tests: comparison of medians (or means?)
In this series of exercises, we will explore how to compare two groups of data, using non-parametric tests (or parametric tests… if it makes sense to do so).

## 2.1 Make clear what is a statistical test, and when to use it.
**Below are some questions to guide the analysis.**

For this part, i used an LLM (gpt 5) because some of these concepts are universal, non-controversial, and always defined the same way across sources. 

Prompt used:
“Give me short, clear, for the following: What is a mean? What is a median? [...] Each answer should be concise, one to two sentences.”

- Mean: The mean is the arithmetic average, obtained by summing all observations and dividing by the number of values.
- Median: The median is the middle value of an ordered dataset, with half of the values below and half above.
- Variance: Variance measures how spread out the data are by averaging the squared deviations from the mean.
- Standard deviation: The standard deviation is the square root of the variance, providing a measure of spread in the same units as the data.
 - Statistical test: A statistical test is a procedure used to evaluate a hypothesis, determining whether observed results are compatible with a null hypothesis.
 - When to use a statistical test: A statistical test is used when we want to know whether an observed difference or association is likely due to chance or reflects a real effect.
- Paired vs unpaired tests: A paired test compares two related measurements from the same subjects (e.g., before/after), while an unpaired test compares two independent groups.


What is a parametric test? A non-parametric test?
What are the assumptions of parametric tests?
What are the assumptions of non-parametric tests?
What is a p-value?
What is the risk of error when using a statistical test?
What is the difference between a paired and an unpaired test?
You can use your favorite LLM, book or web site to answer these questions, but you have to provide at least one reference for each answer (book, article, URL, LLM prompt, etc.).



## 2.2 Effect of treatment over time
This is a very classical question in clinical or sport research: does a treatment-training have an effect over time?

To answer this question scientifically, measurements must be taken before and after treatment, and possibly later, in order to assess whether the effect is lasting or not.

## 2.2.1 The data
Download the file PrePost.csv

This file contains before/after measurements. The treatment is a rehabilitation training for individuals with cardiac conditions.

Before going further, you have to make clear what are the measured variables. What do they measure? What are the units? What are the possible values? What are the expected changes over rehabilitation.

Make a table, with one row per variable, and the following columns:

| Variable | Description | Unit | Possible values | Expected change |

## 2.2.2 The analysis
You want to know if the treatment has an effect on one or more of the measured variables.

Does the treatment have an effect?

What comparison are you going to make?

Which test will you use? Parametric or non-parametric?

Which graph illustrates the effect of the treatment?

What sentence will you write in your thesis to explain your result?

## 2.3 Testing some stereotypes
Humans have stereotypes, some of them are probably true, some others are probably false (e.g., ref ).

Here, we will test stereotypes about snorers… among others.

## 2.3.1 The data
Download the file snore.txt

This file contains anthropometric and qualitative measurements (1 person per line).

2.3.2 The analysis
Do the data confirm the following stereotypes? Provide a reasoned answer (data, figure, result sentence) for each question.
Are snorers fatter?
Do snorers drink or smoke more?
Are men fatter?
Do women smoke less?
From a more general perspective:
Are there any correlations between variables?
3 Epilogue: beyond the scope of this series
Finally, some of you may want to go beyond the scope of this series with new questions, such as:

Can you predict the value of one variable from the others? (linear regression, logistic regression, etc.)
Can you classify individuals based on their variables? (clustering, PCA, etc.)
NOTE: These questions are more advanced, and I don’t want you to address them from a technical standpoint or provide code. But they give you an idea of what you can do with data mining techniques… provided you know why you want to do it. I invite you to think about the perspective these questions open up and, if you feel ready to do so, to write a short paragraph about it.
